\section*{\huge Acknowledgements}

A year at Oxford felt like an impossibility not too long ago. Yet here I am in the closing days of my degree---about to finish the impossible. These efforts are a product of the time and patience others have had with me, who have nurtured my thoughts, and inspired me to seek opportunity to learn. These opportunities, occuring over the past few years and culminating in this work, have brought me to the strangest corners of human thought. I am content to have made it so far, and still so eager to see more of what's out there.

My friends made this journey far easier. To venture, somewhat alone, into these odd corners of human thought can make one feel alone in their thoughts. Yet the ear of a friend goes a long way. And at times, they renew interest in lost thoughts or open new avenues to explore. Their stories and perspectives have helped my view of the world grow more complete, both in and outside of research. I'm grateful.

My advisers made this journey far easier. They too have spent time in odd places, far more time than me. But it's this familiarity with the terrain, the heuristics of thought they've developed, which have made traversing my research more comforting. They've been an ear to some of my wild ideas, and have supported me through a balance of guidance and intellectual freedom. They've nudged me into better directions when I've been lost, and at the same time let me run with whatever's going on in my head. To conceive and to execute on such odd ideas with them has been an experience I cherish.

